---
title: "Assignment B1 - Andrea Wong Koo"
output: github_document
date: "2024-10-29"
---

``` {r}

library(tidyverse)
library(ggplot2)
library(dplyr)
library(testthat)

#Data packages for examples and tests:
library(survival)
library(palmerpenguins)
```


# Exercise 1: Make a Function and Exercise 2: Document your Function

The following is the code for my created function `summarize_category_count()`, with `roxygen2` tags for function documentation.

_Extra coding explanations:_ `deparse(substitute())` was used to convert variable name into character string for verifying data class.

``` {r}

#' Title: Function for Summarizing Category Instances (summarize_category_count())
#' This function counts instances of a specific category within a categorical variable, grouped by a group variable. It supports both factor and discrete numeric variable types, and displays the results in a summary table. 
#' 
#' @param data An input data frame containing the variables to analyze.
#' @param group_var The grouping variable to summarize counts by. Named this way to clarify its role in grouping.
#' @param cat_var The categorical variable of interest. This name is a concise reference to "category variable".
#' @param cat The specific category value to be counted. This parameter directly refers to the target category.
#' @param na.rm A reusable variable indicating whether to ignore NA values in the data. Defined globally to allow flexibility in handling missing values, while reducing redundancy.
#' @param .groups A reusable variable to control the behavior of grouping (e.g., "drop"), as in `dplyr::summarize()`. Defined globally to allow flexibility and reduce redundancy if changing its assignment.
#'
#' @return A summary table with the counts for the chosen category per group.
#' @export
#'
#' @examples
#' 

summarize_category_count <- function(data, group_var, cat_var, cat, na.rm = TRUE, .groups = "drop") {
  
  # To capture the names of the variables for easier use in the plot: 
  group_var_name <- deparse(substitute(group_var)) 
  cat_var_name <- deparse(substitute(cat_var))

  cat_var_data <- data[[cat_var_name]] # Accessing variable from the data
  
  # To ensure that the input variables are part of the dataset:  
  if (!(group_var_name %in% names(data))) {
    stop(paste("Error: The grouping variable", group_var_name, "is not a column in the dataset.")) 
  }
  
  if (!(cat_var_name %in% names(data))) {
    stop(paste("Error: The category variable", cat_var_name, "is not a column in the dataset."))
  }

  # To ensure that the category variable is only a factor or integer numeric type:
  if (!is.factor(cat_var_data) && 
      !(is.numeric(cat_var_data) && 
        all(cat_var_data == as.integer(cat_var_data), na.rm = na.rm))) {
    stop("Sorry, this function only works for factor or discrete numeric (integer) variables. You have provided an object of class: ", 
         class(cat_var_data)[1], ", or a non-integer numeric type.") 
  }
    
  # To group by the specified variable and count instances of the category: 
  summary_data <- data %>% 
    group_by({{group_var}}) %>% 
    summarize(count = sum({{cat_var}} == cat, na.rm = na.rm), .groups = .groups)
    
  print(summary_data) # To print a summary result table. 
  
}

```

# Exercise 3: Include examples

### Examples with `lung` data: 

To use an example with a data set containing a categorical variable with discrete numbers as categories, we can use the `lung` dataset from the `survival` package.

``` {r}
#To explore the `lung` dataframe variables and its documentation: 
glimpse(lung)
?lung
```

``` {r}
#To see a preview of the `lung` data:
head(lung)

```


**Example 1:** An example that counts the instances of an ECOG performance rating (`ph.ecog`) by the sex group variable (`sex`). 

For the group variable, `sex` is either 1 (Male) or 2 (Female).
For the category variable input, the possible `ph.ecog` categories are the following: 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed <50% of the day, 3= in bed > 50% of the day but not bedbound, 4 = bedbound. As seen in `?lung` and `glimpse()` above, the categorical variable (`ph.ecog`) is an integer numeric (`dbl`subtype) class, where each number corresponds to a rating category. 
``` {r}
# Example that counts the instances of ECOG rating category '0' (asymptomatic) by sex: 
summarize_category_count(data = lung, group_var = sex, cat_var = ph.ecog, cat = 0) 
```

**Example 2:** An example that counts the instances of a censoring status category (`status`) by the sex group variable (`sex`). 

For the group variable, `sex` is either 1 (Male) or 2 (Female).
For the category variable input, the `status` can be 1 (censored) or 2(dead). As seen in `?lung` and `glimpse()` above, the categorical variable (`status`) is an integer numeric (`dbl`subtype) class, where each number corresponds to a status category. 
``` {r}
# Example that counts the instances of status category '2' (dead) by sex: 
summarize_category_count(data = lung, group_var = sex, cat_var = status, cat = 2) 

```


### Example with `mtcars` data: 

To create examples using another dataset, we can use the `mtcars` dataframe.

``` {r}
#To explore the `mtcars` dataframe variables and its documentation: 
glimpse(mtcars)
?mtcars
```

``` {r}
#To see a preview of the `lung` data:
head(mtcars)

```

**Example 3:** An example that counts the instances of an engine type category (`vs`) by the transmission group variable (`am`). 

For the group variable, transmission (`am`) is either 0 (automatic) or 2 (manual).
For the categorical variable, the engine type (`vs`) can either be 0 (V-shaped) or 1 (straight).  As seen in `?mtcars` and `glimpse()` above, the categorical variable (`vs`) is an integer numeric (`dbl`subtype) class, where each number corresponds to a status category.

``` {r}
# Example that counts the instances of engine category '1' (straight) by transmission type:
summarize_category_count(data = mtcars, group_var = am, cat_var = vs, cat = 1)  
```

### Example with `penguins` data: 

Another dataset we can use is the `mtcars` dataframe.

``` {r}
#To explore the `penguins` dataframe variables and its documentation: 
glimpse(penguins)
?penguins
```

``` {r}
#To see a preview of the `penguins` data:
head(penguins)

```

**Example 4:** An example that counts the instances of an species type category (`species`) by the sex group variable (`sex`). 

For the group variable, sex (`sex`) is either female or male.
For the categorical variable, the species (`species`) can either be Adélie, Chinstrap and Gentoo. As seen in `?penguins` and `glimpse()` above, the categorical variable (`species`) is factor type, with different levels as categories.

``` {r}
# Example using factor as the categorical variable, counting the instances of the species category "Gentoo" by sex: 
summarize_category_count(data = penguins, group_var = sex, cat_var = species, cat = "Gentoo")

```


# Exercise 4: Test the Function

For the test for handling NA values in the data, here is a synthesized tibble to be used later on: 
``` {r}
modified_lung <- tibble(
  status = c(1, 2, 1, 2, 1, 2, 1, 2, NA, 1),  # Status (1 = alive, 2 = dead)
  age = c(71, 68, 65, 74, 58, 62, 77, 60, 55, 80),  # Age in years
  sex = c(1, 2, 1, 2, 1, 2, 1, NA, 1, 2),  # Sex (1 = male, 2 = female)
  ph.ecog = c(1, 2, NA, 0, 1, 3, 2, 0, NA, 4)  # ECOG performance scores
)

```

The following are test cases to validate the behaviour of the function.
``` {r}
test_that("Testing category summarization function", {
  # Test for the lung dataset with expected category count
  expect_equal(
    summarize_category_count(data = penguins, group_var = sex, cat_var = species, cat = "Gentoo"), 
    penguins %>%
      group_by(sex) %>%
      summarize(count = sum(species == "Gentoo", na.rm = TRUE), .groups = "drop")
  )
  
  # Test for the starwars dataset expecting an error for the wrong input type (decimal instead of integer):
 expect_error(
        summarize_category_count(data = starwars, group_var = sex, cat_var = birth_year, cat = 41.9), 
        regexp = "Sorry, this function only works for factor or discrete numeric \\(integer\\) variables\\. You have provided an object of class: numeric, or a non-integer numeric type."
    )
 
  # Expects the count to match the summarized data while handling NA values
  expect_equal(
    summarize_category_count(data = modified_lung, group_var = sex, cat_var = ph.ecog, cat = 1, na.rm = TRUE), 
    modified_lung %>%
      group_by(sex) %>%
      summarize(count = sum(ph.ecog == 1, na.rm = TRUE), .groups = "drop")
  ) 
  
  # Expects an error when an input category name is not a column in the data set:
  expect_error(
    summarize_category_count(data = mtcars, group_var = transmission, cat_var = vs, cat = 1),
    regexp = paste("Error: The grouping variable", "transmission", "is not a column in the dataset."))
})



```




